---
title: "PA1_template.Rmd"
author: "Caio Alves"
date: "9/18/2018"
output: md_document
---
```{r load_libs, include=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)
options(warn=-1) # Supressing warnings in this report
```
# Peer-graded Assignment: Course Project 1

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device.

## Observations
* I've hide all library loads from HTML/Markdown report for reading purposes
* 

# Analysis
## Loading and preprocessing the data
### Load the Data
```{r load_data}
rawData <- read.csv("activity.csv")
```
### Process/transform the data (if necessary) into a format suitable for your analysis
Creating variable without NA steps
```{r process_transform}
# Using dplyr library
data <- rawData %>% filter(!is.na(steps))
```
## What is mean total number of steps taken per day?
For this part of the assignment, I will ignore the missing values in the dataset.

### Calculate the total number of steps taken per day
```{r total_steps}
totalSteps <- aggregate(data$steps, list(date = data$date), sum)
print(totalSteps)
```

### Make a histogram of the total number of steps taken each day
```{r histogram}
histogram <- ggplot(data = totalSteps, aes(totalSteps$x)) + geom_histogram() + labs(x = "Steps per day", title = "Histogram for steps") + scale_y_continuous(breaks = (1:5)*2)
print(histogram)
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r mean_median_steps}
meanSteps <- aggregate(data$steps, list(date = data$date), mean)
print(meanSteps)
medianSteps <- aggregate(data$steps, list(date = data$date), median)
print(medianSteps)
```

## What is the average daily activity pattern?

### Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r time_series}
timeSeriesData <- aggregate(data$steps, list(interval = data$interval), mean)
timeSeries <- ggplot(data = timeSeriesData, aes(x = interval, y = x)) + geom_line() + labs(x = "5-minute interval", y = "Average Steps", title = "Average steps time series")
print(timeSeries)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max_steps_interval}
maxStepsInteval <- timeSeriesData %>% filter(x == max(x))
print(maxStepsInteval)
```

## Imputing missing values

### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r NA_val}
# There are no NA values for 'date' or 'interval'
nrow(rawData %>% filter(is.na(steps)))
```

### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r strategy}
# The 'steps' median value is 0, as found before.
# My strategy will fill NA values with 0

newDataset <- rawData
newDataset$steps[is.na(newDataset$steps)] <- 0
nrow(newDataset %>% filter(is.na(steps)))
```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r histogram_mean_median_steps}
totalSteps2 <- aggregate(newDataset$steps, list(date = newDataset$date), sum)
histogram2 <- ggplot(data = totalSteps2, aes(totalSteps2$x)) + geom_histogram() + labs(x = "Steps per day", title = "Histogram for steps") + scale_y_continuous(breaks = (1:5)*2)
print(histogram2)

meanSteps2 <- aggregate(newDataset$steps, list(date = newDataset$date), mean)
print(meanSteps2)
medianSteps2 <- aggregate(newDataset$steps, list(date = newDataset$date), median)
print(medianSteps2)

# Q: Do these values differ from the estimates from the first part of the assignment? 
# A: The histogram shows more values for 0 steps; dates previously missing appear now with mean and median values

# Q: What is the impact of imputing missing data on the estimates of the total daily number of steps?
# A: In this case, since I replaced NA with 0, it has no impact
```

## Are there differences in activity patterns between weekdays and weekends?

### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r weekdays}
# Will ignore missing values for this step

weekDataset <- data
weekDataset$weekday <- ifelse(weekdays(as.Date(weekDataset$date)) %in% c("Saturday", "Sunday"), "weekend", "weekday")
```

### Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r weekdays_plot}
weekdayDataset <- weekDataset %>% filter(weekday == "weekday")
timeSeriesData2 <- aggregate(weekdayDataset$steps, list(interval = weekdayDataset$interval), mean)
timeSeries2 <- ggplot(data = timeSeriesData2, aes(x = interval, y = x)) + geom_line() + labs(x = "5-minute interval", y = "Average Steps", title = "Average steps time series - Weekdays")
print(timeSeries2)

weekendDataset <- weekDataset %>% filter(weekday != "weekday")
timeSeriesData3 <- aggregate(weekendDataset$steps, list(interval = weekendDataset$interval), mean)
timeSeries3 <- ggplot(data = timeSeriesData3, aes(x = interval, y = x)) + geom_line() + labs(x = "5-minute interval", y = "Average Steps", title = "Average steps time series - Weekends")
print(timeSeries3)
```
